Dynamics of thin liquid film flows studied with the lattice Boltzmann method

### 2023-07-24 chapters/Theory.tex
Chapter two feedback Jens

Motivated in the Chapter~\ref{chapter:intro} the starting point to study fluid flows are the Navier-Stokes equations~\cite{Navier, Stokes}.
Claude Navier in 1823 as well as Gorge G. Stokes in 1842 derived the governing equations for a fluid flow.

\section{Thin liquid films}
\label{sec:thin_films}
\begin{equation}\label{eq:simple_thin_film}
    \partial_t h = -\partial_x\left(Q(h)\gamma\partial_x^2 p\right),
\end{equation}

\section{From Shallow Water to Thin Liquid Films}
\label{sec:theory_thin_film}

>>>>

The starting point to study fluid dynamics is the Navier-Stokes equation~\cite{claude-louis-marie-henriMemoireLoisMouvement1827, stokesSteadyMotionIncompressible1848}, as motivated in Chap.~\ref{chapter:intro}.
Both Claude Navier in 1823 as well as Gorge G. Stokes in 1842 derived the momentum equation for a Newtonian liquid.
Historically these two were not the only ones. 
Siméon Denis Poisson and Barré de Saint-Venant came to the same conclusion around the same time~\cite{poissonNouvelleTheorieAction1831, de1843notea, jrHistoryAerodynamicsIts1998}.
But history is written and Navier as well as Stokes are identified with the derivation of the momentum of a fluid element.

\section{The Navier-Stokes equation}
\label{sec:navier_stokes_sec}
Mathematically the Navier-Stokes equation can be derived from one definition and two constraints.
The two constraints are mass and momentum conservation.
Mass conservation simply states that mass cannot be lost or created.
The same argument holds for the momentum.
By Noethers theorem every conserved quantity yields a symmetry and vice versa.
In this case conservation of momentum allows us to use the constraints on the solutions that translational symmetry requires.
Put simply a sliding drop experiment has to yield the same result in any laboratory on earths surface, independent of the exact location of the laboratory.
The last ingredient for the derivation is the definition of a fluid.
A fluid is ``\textit{a substance that flows and is not solid}`` according to the Cambridge Dictionary.
This definition can be refined, because a fluid is either a gas or a liquid that continuously deforms under shear stress, at least in the case of a Newtonian fluid.
And as such there is a linear relation between shear rate and shear velocity for a Newtonian fluid.
Clearly there are other liquids (which encloses granular systems to some extent) that exhibit different behaviour, for example toothpaste.
As long as there is no force applied to the tube the toothpaste does not flow.
Even the acceleration due gravity when pointing the tube towards the ground does not exert enough force to generate a flow.
In contrast to Newtonian liquids those liquids have a different response function to shear, they can be addressed as \textit{power law} fluids.
While the phenomenology of these fluids, e.g. shear thinning or thickening, is very interesting and important in many industrial applications it is beyond the scope of this work. 
That said, in Chapter~\ref{chapter:second_paper} an extension towards more complex flows is presented. 
Instead of however changing the shear response we model thermal fluctuations to approximate the stochastic thin film equation (STF).
In fact, the step to derive an extension to our approach are nicely summarized in Chapter~\ref{chapter:second_paper}. 
It should serve as a guideline for further features, such as viscoelasticity for non-Newtonian liquids~\cite{bouchutNewModelShallow2013}.

Knowing what a fluid is and taking into account the conservation of mass and momentum it is possible to derive the Navier-Stokes equation. 
By applying a continuity equation we assume that the fluid itself is a continuum and therefore ignoring the underlying molecular structure for now. 
This allows us to describe the fluids state with a density $\rho(\mathbf{x},t)$, instead of describing every atom or molecule individually.
Assuming an infinitesimal small volume of fluid $V_0$ its density can only change if there is a flow through the boundaries of the volume $\partial V_0$,
\begin{equation}\label{eq:cont_1}
    \partial_t\int_{V_0}\rho\diff V = -\oint_{\partial V_0} \rho\vec{u}\cdot\diff \vec{A}, 
\end{equation}
where $\vec{u}(\vec{x},t)$ is the fluid's velocity, A is the area that encloses the volume $V_0$ and $\diff\vec{A}$ are the normals pointing outwards of the volume.
Violations of this Eq.~(\ref{eq:cont_1}) are equivalent to generation or destruction of mass.

Thanks to Gauss the integral over a closed surface can be written as divergence of a volume and thus the above equation becomes~\cite{koniglichegesellschaftderwissenschaftenzugottingenCommentationesSocietatisRegiae1811}
\begin{equation}\label{eq:cont_2}
    \partial_t\int_{V_0}\rho\diff V = -\int_{V_0} \vec{\nabla}\cdot(\rho\vec{u}) \diff V. 
\end{equation}
Since both sides are integrals over the volume $V_0$ we drop them and work with the differential equation
\begin{equation}\label{eq:cont_3}
    \partial_t\rho + \vec{\nabla}\cdot(\rho\vec{u}) = 0,
\end{equation}
which will be addressed as the continuity equation throughout the thesis.
Actually one finds this equation not only in fluid dynamics for mass conservation but as well in electrodynamics for charge conservation, where the second term is often referred to as flux~\cite{jacksonClassicalElectrodynamics2021, griffithsIntroductionElectrodynamics2013}.

For the momentum equation the same line of arguments is applied, however momentum can change for different reasons.
Momentum can for example be advected with the flow from or into the volume $V_0$.
Pressure on the other hand can be a source of momentum change and of course forces that are applied to the fluid are as well sources~\cite{krugerLatticeBoltzmannMethod2017}.
Collecting these terms and writing down an equation for the momentum we have,
\begin{equation}\label{eq:navier_stokes_1}
    \partial_t \int_{V_0}\rho\vec{u}\diff V = -\oint_{\partial V_0}(\rho\vec{u}\vec{u})\cdot\diff\vec{A} - \oint_{\partial V_0} p\diff\vec{A} + \int_{V_0} \vec{F}\diff V, 
\end{equation}
where $\vec{u}\vec{u}$ is used to denote the outer product
\begin{equation}
    \mathbf{u}\mathbf{u} = \begin{bmatrix}
    u_x u_x & u_x u_y & u_x u_z \\
    u_y u_x & u_y u_y & u_y u_z \\
    u_z u_x & u_z u_y & u_z u_z
    \end{bmatrix} .
\end{equation}
Using the divergence theorem again to reformulate everything in terms of volume integrals and dropping the integration the above equation becomes
\begin{equation}\label{eq:navier_stokes_2}
    \partial_t(\rho\vec{u}) = -\vec{\nabla}\cdot(\rho\vec{u}\vec{u}) - \nabla p + \vec{F}.
\end{equation}
This however is not yet the Navier-Stokes equation, in fact this is the Euler equation~\cite{batchelorIntroductionFluidDynamics1967}.
In the derivation there is one term missing, because the momentum of a fluid can also change due to internal friction, or put simply, due to viscosity.
These viscous contributions are collected in the viscous stress tensor $\hat{\sigma}$ as
\begin{equation}\label{eq:stress_tens}
    \hat{\sigma}_{ij} = \mu\left(\frac{\partial u_i}{\partial x_j} + \frac{\partial u_j}{\partial x_i}\right) + \lambda\delta_{ij}\frac{\partial u_k}{\partial x_k},
\end{equation}
This tensor introduces a lot of the complexity of the Navier-Stokes equation as in principle neither $\mu$ nor $\lambda$ need to be constants and in fact they are tensors themselves.

For the sake of clarity however here it is assumed that both $\mu$ and $\lambda$ are constants.
Inclusion of pressure to $\hat{\sigma}$ yields the total stress tensor $\tilde{\sigma}$. 
Due to symmetries of the problem there are various constraints on this tensor.
A fluid element can be rotated similar to a solid body, but it can also be stretched. 
In contrast to a solid body a fluid element is allowed to deform.
Working out these constraints we find that the pressure has to be a diagonal term and thus
\begin{equation}\label{eq:total_stress}
    \tilde{\sigma}_{ij} = - p \delta_{ij} + \mu\left(\frac{\partial u_i}{\partial x_j} + \frac{\partial u_j}{\partial x_i}\right) + \lambda\delta_{ij}\frac{\partial u_k}{\partial x_k}.
\end{equation}
Now to get from the Euler equation (Eq.~(\ref{eq:navier_stokes_2})) to the Navier-Stoke we add the viscous stress tensor and rearrange terms,
\begin{equation}\label{eq:navier_stokes_3}
    \partial_t(\rho\vec{u}) + \vec{\nabla}\cdot(\rho\vec{u}\vec{u}) = - \vec{\nabla} p + \vec{\nabla}\left[\mu(\vec{\nabla}\vec{u} + \vec{\nabla}\vec{u}^T) + \left(\eta-\frac{2\mu}{3}\right)\vec{\nabla}\cdot\vec{u}\right] + \vec{F},
\end{equation}
where we have used the differential form of Eq.~(\ref{eq:total_stress}) and $\lambda = \eta - 2/3\mu$.

Instead of trying to solve this straight away, let's first discuss what we try to archive.
The main goal of this thesis is to model the dynamics of thin film flows.
An area of fluid dynamics usually addressed as creeping flows. 
The flow velocities in this regime are in fact much slower than the speed of sound $c_s$ ($c_s^{\ce{H2O}} \approx 1500m/s $). 
This allows for the simplification of an incompressible medium.
For such media or fluids one can assume that $\rho = const.$ and therefore by Eq.~(\ref{eq:cont_3}) 
\begin{equation}\label{eq:incomp}
    \vec{\nabla}\cdot\vec{u} = 0.
\end{equation}
A note of caution: While a constant density ($\rho = const.$) does satisfy the incompressibility constraint it is not unique to the underlying thermodynamics.
The thermodynamic compressibility $\kappa$ measures the change of volume with pressure and is given by 
\begin{equation}
    \kappa = \left(\frac{\partial V}{\partial p}\right)_T.
\end{equation}
Strictly speaking a fluid is incompressible as long as its density does not depend on the pressure $p$ or $\kappa = 0$.

Inserting this constraint Eq.~(\ref{eq:incomp}) into Eq.~(\ref{eq:navier_stokes_3}) yields therefore the Navier-Stokes equation for an incompressible fluid as
\begin{equation}\label{eq:navier_stokes_4}
    \rho(\partial_t\vec{u} + \vec{\nabla}\cdot\vec{u}\vec{u}) = - \vec{\nabla} p + \mu\Delta\vec{u} + \vec{F},
\end{equation}
where the symbol $\Delta$ denotes the laplace operator ($\Delta f = \vec{\nabla}\cdot\vec{\nabla} f$).
The left hand side can be rewritten using the material derivative
\begin{equation}\label{eq:mat_dev}
    D_t = \partial_t + \vec{u}\cdot\vec{\nabla}, 
\end{equation}
which is a convenient measure for following a fluid element instead of a single point.
Inserting this, Eq.~(\ref{eq:navier_stokes_4}) can be written as
\begin{equation}\label{eq:navier_stokes_fin}
    \rho D_t \vec{u} =  - \vec{\nabla} p + \mu\Delta\vec{u} + \vec{F}.
\end{equation}
Eqs.~(\ref{eq:cont_3}, \ref{eq:navier_stokes_4}) cannot be solved because there are too few equations for too many unknowns.
So far there are five unknowns ($\rho, \mathbf{u}, p$) for four equations.
The clever way out of this misery is to use an equation of state that relates pressure, density, temperature, energy and entropy.
Especially the relation between pressure, density and temperature helps in closing the system of equations.
To define an equation of state is far from trivial.
It requires a thermodynamically consistent approach that is based on observations or theoretical assumptions.
The equation of state for an ideal gas was born from theoretical assumptions, e.g. non-interacting gas molecules, and reads~\cite{schwablStatisticalMechanics2006}
\begin{equation}\label{eq:eq_of_state}
    p = \rho R T,
\end{equation}
where $R$ is the specific gas constant and $T$ is the temperature.
Having a well defined state in the thermodynamical sense allows to relate pressure and density and therefore to close the system of equations.
The equation of state for an ideal gas, see Eq.~(\ref{eq:eq_of_state}), will be revisited in Chapter~\ref{chapter:method} where we discuss the lattice Boltzmann method.

The Navier-Stokes equation, see Eqs.~(\ref{eq:cont_3}, \ref{eq:navier_stokes_4}), offers a general framework to deal with fluid dynamics problems.
In the case of an incompressible non turbulent flow we have a solid understanding of the velocity field.
For compressible flows with developed turbulence on the other hand we still miss fundamental knowledge, as mentioned in Chapter~\ref{chapter:intro}.
This missing information is the reason that the automotive as well as aviation industry invests heavily in both real world experiments and CFD. 
For the concrete example of thin film flows, however, there is no need to ``solve'' the Navier-Stokes equation.
That is why in the next two sections different theories that build upon the Navier-Stokes are introduced.
Both of these theories using a scaling argument to severely simplify the resulting system of equations.

\section{Shallow Water Theory}
\label{sec:theory_shallow_water}
Solving the Navier-Stokes equation for the problems discussed in Chapters~\ref{chapter:fourth_paper}-\ref{chapter:third_paper} is not needed.
Looking at Figs.~\ref{fig:examples_intro}-\ref{fig:morph_transition} one can identify that the height or thickness of the fluid layer is much smaller than the horizontal extension of the substrate.
This statement together with some more approximations allows to derive another system from the Navier-Stokes equation and the continuity equation which is called the shallow water or Saint-Venant equations~\cite{bTheorieMouvementNonpermanent1871}.

\subsection{Definition}
\begin{figure}
    \centering
    \includegraphics[width=0.55\textwidth]{graphics/simple_shallow_water.pdf}
    \caption{Schematic description of the shallow water problem. A liquid column with position of the free surface $\eta(x,y,t)$ covers a bottom topography $b(x,y)$.
    The liquid's height with the addition of the elevation of the topography defines the resulting height or z-component of the free surface.}
    \label{fig:shallow_water_drawing}
\end{figure}

The shallow water equations are a well working approximation to describe \textit{shallow} flows~\cite{tanShallowWaterHydrodynamics1992} and illustrated in Fig.~\ref{fig:shallow_water_drawing}.
``Shallow'' by definition of the Cambridge dictionary means ``\textit{having only a short distance from the top to the bottom}``, which in fact describes quite well what the model does.
Instead of having a fully resolved three-dimensional model of the fluid's density and velocity the shallow water approach assumes that the degrees of freedom in the vertical direction can be integrated out. 
Therefore all dynamics of the system are effectively captured by the evolution based on the horizontal degrees of freedom $(x,y)$.

Such strong assumptions require to take a look at the system at hand.
Coming back to the meaning of the word shallow, a shallow flow with a liquid column of height $h$ and horizontal length scale $L$ needs to satisfy
\begin{equation}\label{eq:thickness_length}
    \varepsilon := \frac{h}{L} \ll 1,
\end{equation}
where $\varepsilon$ is a measure for the difference between horizontal and vertical scales.
An illustrative example that satisfies this inequality is any of the three major oceans.
On average the oceans are about $3000m$ deep. 
They cover about $71\%$ of the earth's surface area which is approximately $361000000km^2$.
Dividing this number by three, to separate between Atlantic, Pacific and Indian ocean one gets a square with a side length of $~11000km$.
Inserting numbers in Eq.~(\ref{eq:thickness_length}) yields a ratio $\varepsilon \approx 3\cdot 10^{-4}$.
The average height of the fluid column is negligibly small as compared to its horizontal scales.
Therefore, the dynamics in the vertical direction can be considered instantaneous. 
The oceans are not the only dynamical system that with such a relation, the atmospheric layer also has a small $\varepsilon$. 
Atmospheric dynamics can be described with the shallow water model, at least to some extent.

Secondly comes the assumption of the pressure free surface.
Putting it simply, the shallow water model does not know what happens above the fluid.
The assumption is that the density of the layer above the fluid column is small when compared to the fluid density.
The oceans are mostly filled with water, the density of (salt) water is around a thousand times higher than the density of the air above ($\rho^{\ce{H2O}} \approx 1000$kg/m$^3$, $\rho^{\text{air}} \approx 1,25$kg/m$^3$).
To some extent this is, again, actually true for the atmospheric layer.
Using the barometric formula to relate the density with altitude,
\begin{equation}\label{eq:bareometric_form}
    \rho(h_1) = \rho(h_0)e^{-\frac{\Delta h}{h_s}},
\end{equation}
with $\Delta h = h_1 - h_0$ and $h_s = RT/Mg$ is the scale height.
Of course, the model can be refined with a so called multi layer approach~\cite{tubbsMultilayerShallowWater2009, prestininziEffectIntrinsicViscosity2013}, but by definition the assumption is that the upper end of the fluid column does not feel any pressure from above~\cite{tanShallowWaterHydrodynamics1992, jamesNewFrictionModel2019}.

Third, the horizontal velocity of the fluid has to vanish at the ``solid'' ground, or in terms of Fig.~\ref{fig:shallow_water_drawing} looking down from the free surface at $-h(x,y)$.
Although there is a viscosity associated with the fluid, friction with the bottom topography is the dominating dissipation channel.
This kind of boundary condition has been introduced in Chapter~\ref{chapter:intro} is called \textbf{no-slip} boundary condition~\cite{richardsonNoslipBoundaryCondition1973, degennesFluidWallSlippage2002, salmonLatticeBoltzmannMethod1999}.
While this may seem counterintuitive because fluid has no problem to flow over yet dry surfaces, modelling wetting or the wet-dry transition is a complex task, as seen later in this chapter and in Chapter~\ref{chapter:first_paper}.
After all, a theoretical model which is defined as a set of partial differential equations cannot account for the true complexity of nature~\cite{scrivenPhysicsApplicationsDIP1988}.
We come back to the idea of the no-slip boundary condition in Sec.~\ref{sec:thin_films}.

The big circular flow systems in the oceans like the Gulf Stream have much larger velocity components in the horizontal directions as compared to the vertical one.  
This is actually crucial to know because in the following mathematical modelling it is taken into account and will allow for further reduction of complexity.
We therefore assume that the vertical velocity component is small as compared to the horizontal ones.

The task now is to make, Eq.~(\ref{eq:navier_stokes_4}), aware about these constraints.
Although this sounds straightforward partial differential equations are complex. 
They are one of the topics in mathematics where researchers are satisfied with the irritating result that an equation cannot be solved, an undesired outcome for most Physicists.
Nevertheless, to come up with consistent strategies to solve them one can classify them, e.g. hyperbolic, parabolic, linear, nonlinear, initial value problem or boundary value problem.
There is in fact overlap between those terms, however to keep things simple it is sufficient to talk about initial or boundary value problems.
The former can be solved knowing the initial configuration which then defines the evolution, e.g. the initial coordinates and velocity of a pendulum. 
For the latter ones it is mandatory to know the solution at the boundaries. 
If we know the current temperature distribution on a metallic plate and the heat flux on the boundaries then we are able to calculate the temperature field on that plate at later times.
The shallow water system can be derived from the Navier-Stokes equation using both well defined boundary conditions and system specific approximations.
For example the statement \textit{pressure free surface} translates to the fact that there is a boundary condition requiring $p = 0$ at $z = \eta$. 
Therefore no pressure is acting on liquids interface.

\subsection{Scaling analysis}
\label{subssec:scaling_shallow water}
Starting with the system specific approximations, in Eq.~(\ref{eq:thickness_length}) a so-called scaling parameter $\varepsilon$ has been introduced. 
We use this parameter to rescale the velocities according to
\begin{equation}\label{eq:vel_scaling}
    u_x = \tilde{u}_x,\quad u_y = \tilde{u}_y,\quad u_z = \varepsilon\tilde{u}_z.
\end{equation}
For an incompressible fluid with vanishing density gradient Eq.~(\ref{eq:cont_3}) therefore becomes
\begin{equation}\label{eq:cont_nondim}
    \vec{\nabla}\cdot\vec{u} = \partial_x u_x + \partial_y u_y + \varepsilon\partial_z u_z = 0.
\end{equation}
To satisfy that all terms of this equation are the same order in $\varepsilon$ the coordinates need to be rescaled as well. 
This choice however is not unique and offers two possibilities~\cite{jamesNewFrictionModel2019}.
There is on the one hand the long wave approximation~\cite{oronLongscaleEvolutionThin1997},
\begin{equation}\label{eq:long_wave_scaling}
    x = \frac{\tilde{x}}{\varepsilon},\quad y = \frac{\tilde{y}}{\varepsilon},\quad z = \tilde{z},\quad t = \frac{\tilde{t}}{\varepsilon},
\end{equation}
which again is encountered in Sec.~\ref{subsec:thin_film_scaling} and has the effect of smoothing out spatial and temporal variations.
On the other hand is the thin layer scaling
\begin{equation}\label{eq:scaling_thin_layer}
    x = \tilde{x},\quad y = \tilde{y},\quad z = \varepsilon\tilde{z},\quad t = \tilde{t},
\end{equation}
which corresponds to the classical scaling of boundary layer theory~\cite{milne-thomsonTheoreticalHydrodynamics1996}.

In Eqs.~(\ref{eq:vel_scaling}-\ref{eq:scaling_thin_layer}) there is no distinct difference between both horizontal dimensions $(x,y)$. 
For the sake of clarity and to simplify the derivation of the shallow water model we neglect the $y$-components.
Of course, to study large oceanic systems or Coriolis forces one needs both horizontal dimensions, see e.g.~\cite{dellarShallowWaterEquations2005, marcheDerivationNewTwodimensional2007}, but for the following derivation it is sufficient to use only one.
Furthermore we use Eq.~(\ref{eq:long_wave_scaling}), which represents long wave theory or scaling. 
We make use of this approach once more in the derivation of the thin film equation.
The benefit of this choice is the smoothing effect on variations which makes them more gradual. 
But not only the gradients in horizontal dimensions are multiplied with $\varepsilon^{-1}$.
The time is scaled with $\varepsilon^{-1}$ as well which smears the dynamics.
Inserting Eqs.~(\ref{eq:vel_scaling},\ref{eq:long_wave_scaling}) into Eq.~(\ref{eq:cont_3}) yields
\begin{equation}\label{eq:rescaled_cont}
    \tilde{\nabla}\cdot\tilde{\mathbf{u}} = \partial_{\tilde{x}} \tilde{u}_x + \partial_{\tilde{z}} \tilde{u}_z  = 0,
\end{equation}
where two dimensional vectors, such as $\mathbf{u} = (u_x, u_z)$ are bold while three dimensional ones are denoted with an arrow symbol.

Scaling analysis and renormalization are key concepts in modern physics~\cite{yakhotRenormalizationGroupAnalysis1986, IntroductionQuantumField2018}.
One of the key concepts of this idea are symmetries, and therefore transformations under which a given state does not change.
Fluid dynamics to some extent admits a conformal symmetry.
The governing equations describe the flow through a nanotube as well as the circular currents in oceans~\cite{secchiMassiveRadiusdependentFlow2016, marcheDerivationNewTwodimensional2007}.
Some features however such as the drag on an airfoil do not simply scale with the size of the system.
Therefore while a miniaturised model may create enough lift force, simply scaling up the size is not sufficient to ensure enough lift.  
In the following it will be helpful to introduce dimensionless numbers~\cite{ruzickaDimensionlessNumbers2008,neversFluidMechanicsChemical2019}.
These numbers have a long history in fluid dynamics.
The Reynolds number for example is a measure if a flow is turbulent or not, it depends only on proportions and not on an exact experiment.
With the knowledge of these number comparing experiments is straightforward, even if the experiments differ in many aspects.
The two dimensionless numbers of interest for the shallow water theory are the Reynolds number and the Froud number
\begin{equation}\label{eq:Re_and_Fr}
    Re = \frac{\rho L v_0}{\mu},\quad Fr = \frac{v_0}{\sqrt{g L}},
\end{equation}
where $L$ is a characteristic length scale of the problem, e.g. the mean water depth, $v_0$ and $g$ are a characteristic velocity and the gravitational acceleration ($g = 9.81 m/s^2$), respectively.
A characteristic velocity can be for example the inflow velocity from a boundary, or the speed of a certain wave type such as a tsunami.
Both numbers measure the magnitude of certain effects, in the case of the Reynolds number the inertial forces are compared against viscosity.
The Froud number is a measure to compare kinetic ($v_0$) with potential energy ($g L$).

In the next step both $Fr$ and $Re$ are used to make Eq.~(\ref{eq:navier_stokes_fin}) dimensionless.
For convenience the vector equation is split into two respective equations with gravity being the only force present.
Writing the full equation with all components yields
\begin{align}\label{eq:2d_RNS}
    \partial_t u_x + u_x\partial_x u_x + u_z\partial_z u_x &= -\partial_x p + \frac{1}{Re}\Delta u_x ,\\
    \partial_t u_z + u_z\partial_x u_x + u_z\partial_z u_z &= -\partial_z p - \frac{1}{Fr^2} + \frac{1}{Re}\Delta u_z ,
\end{align}
with the two dimensional Laplacian $\Delta = (\partial_x^2 + \partial_z^2)$.
Using long wave scaling, Eq.~(\ref{eq:long_wave_scaling}) and the rescaled length and time variables (e.g. $\tilde{x}$),
\begin{align}\label{eq:2d_RNS_eps}
    \varepsilon(\partial_{\tilde{t}} \tilde{u}_x + \tilde{u}_x \partial_{\tilde{x}} \tilde{u}_x + \tilde{u}_z \partial_{\tilde{z}} \tilde{u}_x ) &= -\varepsilon\partial_{\tilde{x}} \tilde{p} + \frac{1}{Re}(\varepsilon^2 \partial_{\tilde{x}}^2 \tilde{u}_x + \partial_{\tilde{z}}^2 \tilde{u}_x ) ,\\
    \varepsilon^2(\partial_{\tilde{t}} \tilde{u}_z + \tilde{u}_z\partial_{\tilde{x}} \tilde{u}_x + \tilde{u}_z\partial_{\tilde{z}} \tilde{u}_z) &= -\partial_{\tilde{z}}\tilde{p} - \frac{1}{Fr^2} + \frac{\varepsilon}{Re}(\varepsilon^2 \partial_{\tilde{x}}^2 \tilde{u}_z + \partial_{\tilde{z}}^2 \tilde{u}_z ) .
\end{align}
We now assume that terms with higher orders in $\varepsilon$ are negligible and therefore $O(\varepsilon^2) \rightarrow 0$, collecting only terms with up to $O(\varepsilon)$
\begin{align}\label{eq:2d_RNS_linear}
    \partial_{\tilde{t}} \tilde{u}_x + \tilde{u}_x\partial_{\tilde{x}} \tilde{u}_x + \tilde{u}_z\partial_{\tilde{z}} \tilde{u}_x &= -\partial_{\tilde{x}} \tilde{p} + \frac{\partial_{\tilde{z}}^2\tilde{u}_x}{\varepsilon Re}, \\
    0 &= -\partial_{\tilde{z}}\tilde{p} - \frac{1}{Fr^2} + \frac{\varepsilon\partial_{\tilde{z}}^2 \tilde{u}_z}{Re} .
\end{align}
Interestingly in the first of the two above equations the term associated to the horizontal viscosity is $O(\varepsilon^2)$ and can be dropped.
On the other hand with the assumption of the long wave scaling the second of the above equations yields the hydrostatic pressure, taking only $O(1)$ terms
\begin{equation}\label{eq:hydro_static}
    \partial_{\tilde{z}}\tilde{p} = -\frac{1}{Fr^2},
\end{equation}
which can be readily integrated, given $Fr$ is a constant, using the upper bound at the liquid air interface $\eta(x)$ 
\begin{equation}\label{eq:hydro_static_int}
    \tilde{p} = \frac{1}{Fr^2}(\tilde{\eta} - \tilde{z}).   
\end{equation}
This implies that $\partial_{\tilde{x}}p$ does not depend on $\tilde{z}$ and therefore states that the pressure gradient is conserved along z.

Summarizing the impact of the approximation, on the continuity equation Eq.~(\ref{eq:rescaled_cont}) and Eqs.~(\ref{eq:2d_RNS_linear}-\ref{eq:hydro_static}) and dropping the tilde we get 
\begin{align}
    \partial_x u_x + \partial_z u_z &= 0, \label{eq:cont_sw_euler}\\
    \partial_t u_x + u_x\partial_x u_x + u_z\partial_{z} u_x &= -\partial_x p + \frac{\partial_z^2 u_x}{\varepsilon Re}, \label{eq:mom_sw_euler}\\
    \partial_z p &= -\frac{1}{Fr^2} \left(+\frac{\varepsilon\partial_{\tilde{z}}^2 \tilde{u}_z}{Re}\right), \label{eq:hydrostat_sw_euler}
\end{align}
with the addition of boundary conditions at the liquid air interface ($z = \eta$)
\begin{align}
    \partial_t \eta + u_x\partial_x\eta - u_z &= 0, \label{eq:sw_boundaries_top_1}\\
    p &= 0, \label{eq:sw_boundaries_top_2}\\
    \partial_z u_x &= 0, \label{eq:sw_boundaries_top_3}
\end{align}
as well as the no slip condition at the liquid solid interface ($z = b$)
\begin{align}
    u_x = 0, \label{eq:no-slip_sw1}\\
    u_z = 0. \label{eq:no-slip_sw2}
\end{align}
One interesting limit is the case when $\varepsilon Re \rightarrow \infty$.
In this limit both the continuity equation as well as the boundary conditions are untouched.
The only part that is modified is the momentum equation Eq.~(\ref{eq:mom_sw_euler}),
\begin{equation}
    \partial_t u_x + u_x\partial_x u_x + u_z\partial_{z} u_x = -\partial_x p,
\end{equation}
which describes the hydrostatic Euler equations~\cite{brenierGeneralizedSolutionsHydrostatic2008}.

\subsection{Integration along the vertical dimension}
\label{subsec:int_sw}
The classical approach to obtain the shallow water equations is to integrate the system of Eqs.~(\ref{eq:cont_sw_euler}-\ref{eq:hydrostat_sw_euler}) along the vertical dimension.
Having already neglected the $y$-component, the integration will also integrate out degrees of freedom in $z$.
The resulting system will therefore only vary with a $x$ dependency.
Analogous to Fig.~\ref{fig:shallow_water_drawing} by integration along the vertical direction of the three dimensional density field $\rho$, the height of the fluid column emerges as a dynamic quantity,
\begin{equation}\label{eq:shallow_water_height}
    h(x) = \eta(x) - b(x).    
\end{equation}
Therefore it is a reduction of complexity and effectively enslaves the vertical dynamics to the horizontal one.

In terms of velocity a new variable needs to be defined which accounts for the averaging along the vertical
\begin{equation}\label{eq:sw_discharge}
    hU = \int_{b}^{\eta}u_x\diff z,
\end{equation}
where $h$ is the height (Eq.~(\ref{eq:shallow_water_height})) and $U$ the height averaged velocity. 
The pair $hU$, thus height that is transported with velocity $U$, can also be addressed as discharge $Q$~\cite{salmonLatticeBoltzmannMethod1999}.
Having introduced the height $h$ and the discharge $Q$ the next step is to integrate the continuity and momentum equations, Eqs.~(\ref{eq:cont_sw_euler}, \ref{eq:mom_sw_euler}). 
Starting with the former
\begin{align}
    0 =& \int_b^{\eta} (\partial_x u_x + \partial_z u_z) \diff z, \label{eq:vert_int_cont_1}\\
    =& \int_b^{\eta} (\partial_x u_x \diff z) + u_z|_{z = \eta} - u_z|_{z = b}, \label{eq:vert_int_cont_2}
\end{align}
 where the $z$ dependent velocity can be integrated. 
 For the $x$-component we make use of the integration rule of Leibniz, 
\begin{equation}\label{eq:int_leibniz}
    \partial_x\int_{A(x)}^{B(x)}F(x,y)\diff y = \int_{A(x)}^{B(x)}\frac{\partial F(x,y)}{\partial_x} \diff y + \frac{\diff B(x)}{\diff x}F(x,B(x)) - \frac{\diff A(x)}{\diff x} F(x, A(x)),
\end{equation}
which upon inserting yields
\begin{equation}\label{eq:vert_int_cont_3}
    0 = \partial_x\left(\int_{b}^{\eta} u_x \diff z \right) - u_x|_{z=\eta}\partial_x\eta + u_x|_{z=b}\partial_x b + u_z|_{z = \eta}  - u_z|_{z = b}. 
\end{equation}
Simplifying this equation by making use of Eq.~(\ref{eq:sw_boundaries_top_1}) and Eqs.~(\ref{eq:no-slip_sw1}-\ref{eq:no-slip_sw2}) yields
\begin{equation}\label{eq:vert_int_cont_4}
    0 = \partial_x\left(\int_{b}^{\eta} u_x \diff z \right) + \partial_t \eta,
\end{equation}
where the integral is simply the discharge $hU$. 
Assuming that the river or shallow water bed $b(x)$ is time independent the equation can be condensed to
\begin{equation}\label{eq:shallow_water_cont_true}
    \partial_t h + \partial_x(hU) = 0.
\end{equation} 
This is equal to the statement whenever height changes it has to flow somewhere and therefore ensures mass conservation. 
Rivers such as the Colorado River change their bed quite significantly with time, the same is true for coastlines as we know that they erode with time.
The reason we use $b(x)$ and not $b(x, t)$ is the mismatch in time scales between the erosion and the flow.
For example, the Grand Canyon is about 5-6 million years old and roughly 1.9km deep, the Colorado River flows however with a velocity of $\approx 3.5$km/h~\cite{grafColoradoRiverGrand1997}.

Performing the horizontal integration to the depth averaged momentum equation, see Eq.~(\ref{eq:mom_sw_euler}) we have
\begin{align}\label{eq:mom_shallow_water_int}
    \int_b^{\eta}\left(-\partial_x p + \frac{\partial_z^2 u_x}{\varepsilon Re}\right)\diff z &= \int_b^{\eta}\left(\partial_t u_x + u_x\partial_x u_x + u_z\partial_{z} u_x\right)\diff z. 
\end{align}
The two terms on the left hand side can be integrated straight away. 
We know that the pressure gradient $\partial_x p$ is conserved along the vertical dimension. 
The other term ($\propto\partial^2 u$) has a similar shape as the viscous term in Eq.~(\ref{eq:navier_stokes_4}). 
Performing the integral therefore yields
\begin{equation}\label{eq:sw_mom_left}
    \int_b^{\eta}\left(-\partial_x p + \frac{\partial_z^2 u_x}{\varepsilon Re}\right)\diff z = -h\partial_x p + \frac{\partial_z u_x|_{z = \eta} - \partial_z u_x|_{z=b}}{\varepsilon Re},
\end{equation}
with the boundary condition that fluid does not accelerate into the air phase, see Eq.~(\ref{eq:sw_boundaries_top_3}), the left hand side reduces to
\begin{equation}
    \int_b^{\eta}\left(-\partial_x p + \frac{\partial_z^2 u_x}{\varepsilon Re}\right)\diff z = -h\partial_x p - \frac{\partial_z u_x|_{z=b}}{\varepsilon Re}.
\end{equation}
The right hand side of Eq.~(\ref{eq:mom_shallow_water_int}) reads
\begin{align}\label{eq:mom_shallow_rhs}
    \int_b^{\eta}\left(\partial_t u_x + u_x\partial_x u_x + u_z\partial_{z} u_x\right)\diff z &= \int_b^{\eta}\left(\partial_t u_x\right)\diff z \nonumber\\
    &+ \int_b^{\eta}\left(u_x\partial_x u_x + u_z\partial_{z} u_x\right)\diff z ,
\end{align}
where the integral is split into two separate integrals for convenience.
The time dependent part can be easily integrated using Eq.(~\ref{eq:int_leibniz})
\begin{equation}\label{eq:first_term_rhs_mom_sw}
    \int_b^{\eta}\left(\partial_t u_x\right)\diff z = \partial_t \int_b^{\eta} u_x \diff z - \partial_t\eta u_x|_{z=\eta} - \partial_t b u_x|_{z=b} = \partial_t(hU) - \partial_t\eta u_x|_{z=\eta} - \partial_t b u_x|_{z=b}. 
\end{equation}
The other integral is somewhat more tricky.
First, the $u_z\partial_z u_x$ term is integrated by parts
\begin{equation}
    \int_b^{\eta}u_z(\partial_{z} u_x)\diff z = [u_x u_z]_b^{\eta} - \int_b^{\eta} (\partial_z u_z) u_x \diff z.
\end{equation}
We use this intermediate result, Eq.~(\ref{eq:cont_sw_euler}) and the integration rule of Leibnitz to perform the integral of the second line of Eq.(\ref{eq:mom_shallow_rhs}) 
\begin{align}
